---
title: "Impact of Severe Weather Events"
author: "Alex Robinson"
date: "30 June 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Using data from the U.S. National Oceanic and Atmospheric Administration's
(NOAA) storm database an investigation of the economic and health effects of
severe weather events has been undertaken.

Adding together the number of reported fatalities and injuries to give an 
overall casualty figure showed **tornadoes** as by far the most injurious type of
event.

Adding together the crop and property damage figures to give an overall cost 
figure showed **flooding** to be the most costly event type.

## Data Processing

The data for this study is available here:

https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

and information on the data is available here:

https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

```{r read_data, cache=TRUE}
src.data <- "./repdata%2Fdata%2FStormData.csv.bz2"
if(file.exists(src.data)){
        storm.data <- read.csv(src.data)
}
```

To process the data a small amount of data cleansing was undertaken to standardise
capitalisation where possible and to remove "/" characters from the event type
again to standardise the names.

*This study would benefit from a more thorough data cleanse and standardisation
effort. There remain a number of events that will be miscounted due to 
inconsistent classification.*

Those events with either fatalities or injuries were separated out and given a
total casualty figure by summing the fatalities and injury numbers.

The events where property or crop damage were reported were assigned to a 
separate data.frame and an attempt at providing a total damage cost was made.

This was done by examining the PROPDMG (and PROPDMGEXP) and CROPDMG 
(and CROPDMGEXP) values. The DMG figures were mutliplied by a value based on its
EXP field (K = 1, M = 1000, B = 1000000) to give a value in 1000's of dollars. 
These values were then summed.

*Where the EXP column was not K, M or B the figures were not included in the
calculation.*


``` {r process_data}
## tidy up some of the event type data
storm.data$EVTYPE <- toupper(storm.data$EVTYPE)
storm.data$EVTYPE <- gsub("/", " ", storm.data$EVTYPE)
storm.data$EVTYPE[grep("THUNDER", storm.data$EVTYPE)] <- "THUNDERSTORM"
storm.data$EVTYPE[grep("TSTM", storm.data$EVTYPE)] <- "THUNDERSTORM"
storm.data$EVTYPE[storm.data$EVTYPE == "AVALANCE"] <- "AVALANCHE"
## for more accuracy this data cleanse activity should be extended

## Also standardise the EXP data as much as possible
storm.data$CROPDMGEXP <- toupper(storm.data$CROPDMGEXP)
storm.data$PROPDMGEXP <- toupper(storm.data$PROPDMGEXP)

## extract only those cases where people were injured or killed
casualty.data <- storm.data[storm.data$FATALITIES > 0 | storm.data$INJURIES > 0, ]
casualty.data["CASUALTIES"] <- casualty.data$FATALITIES + casualty.data$INJURIES

## extract those cases where some econommic damage was incurred
expense.data <- storm.data[storm.data$PROPDMG > 0 | storm.data$CROPDMG > 0, ]
expense.data["PROPDMGX"] <- 0
expense.data["CROPDMGX"] <- 0
## Turn the EXP information into a figure for multiplication to 1000's
## so K = 1, M = 1,000, B = 1,000,000 anything else will be codified as 0
expense.data$PROPDMGX[expense.data$PROPDMGEXP == "K"] <- 1
expense.data$PROPDMGX[expense.data$PROPDMGEXP == "M"] <- 1000
expense.data$PROPDMGX[expense.data$PROPDMGEXP == "B"] <- 1000000
expense.data$CROPDMGX[expense.data$CROPDMGEXP == "K"] <- 1
expense.data$CROPDMGX[expense.data$CROPDMGEXP == "M"] <- 1000
expense.data$CROPDMGX[expense.data$CROPDMGEXP == "B"] <- 1000000
## Multiply the damage figures by these multipliers
expense.data["PROPDMGTOT"] <- expense.data$PROPDMG * expense.data$PROPDMGX
expense.data["CROPDMGTOT"] <- expense.data$CROPDMG * expense.data$CROPDMGX
expense.data["DMGTOT"] <- 0
## For the total add the 2 figures together
expense.data$DMGTOT <- expense.data$PROPDMGTOT + expense.data$CROPDMGTOT

```
## Results

### Casualty Data

The event types and the total number of casualties for the 5 most injurious
events are shown below.

Tornadoes account for by far the most reported injuries

```{r casualty_results}
casualty.events <- aggregate(CASUALTIES ~ EVTYPE, data = casualty.data, FUN = sum)

casualty.events <- casualty.events[order(-casualty.events$CASUALTIES),]
names(casualty.events) <- c("Event Type", "Casualties")
casualty.events <- head(casualty.events)
format(casualty.events, big.mark = ",")

```

```{r casualty_plot}
barplot(casualty.events$Casualties, names.arg = casualty.events$`Event Type`,
        cex.names = 0.5, col = "sienna", main = "Total Casualties by Event Type")
```

### Expense Data

The event types and the total cost of damages for the 5 most costly events types 
are shown below.

Flooding accounts for the highest cost of damage to crops and property with
hurricane/typhoon second.

```{r expense_results}
expense.events <- aggregate(DMGTOT ~ EVTYPE, data = expense.data, FUN = sum)

expense.events <- expense.events[order(-expense.events$DMGTOT),]
names(expense.events) <- c("Event Type", "Total Damage in $Ks")
expense.events <- head(expense.events)
format(expense.events, big.mark = ",")

```

```{r expense_plot}
barplot(expense.events$`Total Damage in $Ks`, names.arg = expense.events$`Event Type`,
        cex.names = 0.5, col = "chartreuse4", axes = FALSE)
pts <- pretty(expense.events$`Total Damage in $Ks` / 1000000)
axis(2, at = pretty(expense.events$`Total Damage in $Ks`), labels = pts)
title(ylab = "Total damage in Billions $", main = "Economic Impact of Events")
```
